### 2021-08-25 plos_latex_template.tex
RJ: spotlight rule

The first step towards a diverse and inclusive conference is to recognize that people are diverse, that some groups face discrimination and might be underrepresented in your community and event \cite{timperleyHeMoanaPukepuke2020}.

When thinking about diversity, each one of us can have different points of view about the meaning of the term, depending on where we come from. If you are reading this, you may be here because you already have 
Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity,  geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. While human diversity should be celebrated and respected in every space, the truth is that there are implicit hierarchies along these axes, and some statuses (e.g., cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems are consciously and unconsciously built. Thinking about diversity and inclusion is thinking about counteracting the structures that hold these hierarchies in place.

Think about the groups that are usually unsupported or discriminated against in your community or event, and about the meaning of "diverse" in your context. Imagine the result you would like to see if you succeeded in organizing a diverse conference according to your definition. Does this mean an even gender distribution in your speakers? Does this relate to people of all races -especially black people- being present as coordinators, presenters, and in the audience? Does this mean having LGBTQ+ friendly-spaces or community participation from key geographic regions? This vision should guide and help hold the organizing team accountable along the way.

@@

While you may want to improve representation towards some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a truly inclusive environment means taking care of all the other aspects as well. Having consideration of religious practices, setting specific accommodations for lactating women and child care, having LGBTIQ+-friendly spaces, creating community-only spaces, enforcing the use of pronouns, and treating gender as a non-binary variable, and accomodating are some examples of decisions that can make inclusion real. 
%Representation is an important aspect of diversity and inclusion: seeing people like you, from your country, that speak your language is one of the best way to feel you belong
Some teams gather this vision in a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}) to express the values of the conference and make diversity part of the communication strategy see Rule \ref{rule_communication} and in every decision you make, will let people know that they are seen, respected, and welcome; that this is their space and community too. 
However, it is important to not stop with a statement alone, but to identify specific goals and tasks that ensure compliance with these larger goals, for example, identifying accessible platforms, developing a detailed code of conduct and a team to enforce it, and preparing accessibility guidelines (see other Rules below).

@@

\section{Rule X: Have a strong online component of the conference} 
\label{rule_online}
In-person interaction is priceless; however, it is more expensive for some, even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021}. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipourChangingScientificMeetings2021,ninerBetterWhomLeveling2021, gattrellComparisonCarbonCosts}.

@@

\section{Rule X: Search proactively for the best people the conference's vision}
\label{rule_presenters}
When inviting people as keynote speakers, program committee, session chairs and others, we should always look for the best in their subject. Ensuring diversity in each of these roles is a deliberate process -- if there is not much diversity in the first set of names, it needs to be revised to include people from minoritized groups. If these groups are not showing up in the A-list, it is seldom because of their technical acumen -- it is because we, as organizers, are not used to looking in unfamiliar places to bring in the talent. Many of our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyerNoticeWhoScience2021,sarabipourChangingScientificMeetings2021}. Rather than deter us, this should encourage us to look further to find great people that are not routinely in the spotlight. We need to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard these people until there is ample representation across the diversity spectrum and dimensions.

% So, as long as we start with an inclusive and diverse organizing team (\textbf{Rule 2}), we can aim at finding the right people for *these invited roles*. The regional and local communities are also good sources to tap into. For example, for useR! 2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential attendees, and sponsors. %ast: i think this could come back, removing the specific part about "conference organization" - i edited that.

@@

\section{Rule X: Actively reach out and involve people who have been systematically excluded}
\label{rule_communication}
%this rule name should be more explicit towards communication, "promoting"
If part of the community has been historically discriminated against, one should emphasize that they are particularly welcome in this event and that the organization will make it a safe and inclusive environment. Publish a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}), the code of conduct and accessibility guidelines. Fee waivers, financial support, and ease of application/registration should also be advertised; people who are used to applying and getting rejected for scholarships and waivers may find it relieving that the process will be supportive rather than discriminatory against underrepresented groups. Advertise the conference in multiple languages (rule \ref{rule_language}). Express this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) to let people know that they are seen, respected, and welcome; that this is their space and community too.

>>>>

When thinking about diversity, each one of us can have different points of view about the meaning of the term, depending on where we come from. If you are reading this, you may be here because you already have some experience in Diversity, Inclusion, and Accessibility practices. Or you may want to improve diversity in your conference but do not know how or where to start. This article does not intend to be a comprehensive review of this topic --this would be impossible-- but will give some ideas pointing in key directions.

Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity,  geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. While human diversity should be celebrated and respected in every way, the truth is that there are implicit hierarchies along these axes, and some statuses (e.g., cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems --including conferences-- are consciously and unconsciously built. As a result, most people in the world fall into some category where they are part of a marginalized group --as opposed to a privileged one. The first step towards a diverse and inclusive conference is to recognize that people are diverse, that some groups face discrimination and might be underrepresented in your community and event \cite{timperleyHeMoanaPukepuke2020}.

%Thinking about diversity and inclusion is thinking about counteracting the structures that hold these hierarchies in place.

Think about the groups that are usually unsupported or discriminated against in your community or event, and about the meaning of "diverse" in your context. Imagine the result you would like to see if you succeeded in organizing a diverse conference according to your vision. Does this mean an even gender distribution in your speakers? Does this relate to people of all races -especially black people- being present as coordinators, presenters, and in the audience? Does this mean having LGBTQ+ friendly-spaces or community participation from key geographic regions? This vision should guide and help hold the organizing team accountable along the way.

@@

\section{Create a safe and welcoming environment}

While you may want to improve representation towards some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a truly inclusive environment means taking care of all the other aspects as well. Having consideration of religious practices, setting specific accommodations for lactating women and child care, having LGBTIQ+-friendly spaces, creating community-only spaces, enforcing the use of pronouns, and treating gender as a non-binary variable, are just some examples of decisions that can make inclusion real.

%Representation is an important aspect of diversity and inclusion: seeing people like you, from your country, that speak your language is one of the best way to feel you belong

Some teams gather this vision in a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}) to express the values of the conference and make diversity part of the outreach and communication strategy see Rule \ref{rule_communication}.

However, it is important to not stop with a statement alone, but to identify specific goals and tasks that ensure compliance with these larger goals, for example, identifying accessible platforms, developing a detailed code of conduct and a team to enforce it, and preparing accessibility guidelines (see other Rules below).

@@

\section{Rule X: Have a strong online component of the conference} 
\label{rule_online}
In-person interaction is priceless; however, it is more expensive for some, even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021}. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipourChangingScientificMeetings2021,ninerBetterWhomLeveling2021, gattrellComparisonCarbonCosts2021}.

@@

\section{Rule X: Consciously unbias your spotlight roles}
\label{rule_spotlight}

When choosing or inviting people as keynote speakers, program committee, session chairs and other spotlight roles, it is likely that there will not be much diversity in the first set of names. Many of our biased lists are products of the existing systems that have always privileged some groups of people  \cite{dwyerNoticeWhoScience2021,swartzScienceValueDiversity2019,wongBuildDiversityScience2020,dignazioUnicornsJanitorsNinjas2020}. Rather than deter us, this implicit and systemic bias should encourage us to look further to find great people that are not routinely in the spotlight. 
Ensuring diversity in each of these roles needs to be a deliberate process. We need to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard these people until there is ample representation across the diversity spectrum and dimensions.

Make sure that every selection committee --the committee looking for keynote speakers, the selection committee for abstracts, the prizes and award committees --are also diverse \cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}, and ask them to be aware that everyone has an implicit biases, to recognize them, and try to counteract them. An inclusive and diverse organizing team (\textbf{Rule \ref{rule_organizing_team}}) is already a great starting point to overcome this bias in other roles. The regional and local communities in your field are also good sources to tap into. 
For example, for useR! 2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential presenters, and attendees; some of them even held spaces to help the members of their groups to prepare abstracts for submission.

Furthermore, consider bringing to the spotlight the people that have contributed to your field in more collaborative  ways \cite{cheng2020x+}. Give awards (see Rule \ref{rule_accessibility}) to those who prepared accessible slides and presentations, for instance, to acknowledge being mindful of inclusiveness. Community building, for instance, is challenging and usually unrewarded when compared to publications or software development \cite{acionWhyChooseCommunity2020}; give the people who deliberately took the time to work on their communities the recognition they deserve. Defy the stereotypical criteria for success by acknowledging these community practices. 
However, do not restrict people from marginalized groups or community-builders to talk or work only in issues related to Diversity, Inclusion, and Accessibility. Recognize their areas of expertise and respect their will regarding participation (or not) in community-building events.

@@

\section{Rule X: Actively reach out and involve people who have been systematically excluded}
\label{rule_communication}
%this rule name should be more explicit towards communication, "promoting"
If part of the community has been historically discriminated against, one should emphasize that they are particularly welcome in this event and that the organization will make it a safe and inclusive environment. Publish the diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}), the code of conduct and accessibility guidelines. Fee waivers, financial support, and ease of application/registration should also be advertised; people who fear getting rejected when applying for scholarships and waivers may find it relieving that the process will be supportive rather than discriminatory against underrepresented groups. Advertise the conference in multiple languages (rule \ref{rule_language}). Express this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) to let people know that they are seen, respected, and welcome; that this is their space and community too.
